Sequoidal Categories and Transfinite Games: Towards a Coalgebraic Approach to Linear Logic\footnote{This work was partially supported by [TODO: find the appropriate grant codes]}

### 2017-05-08 round-corner/paper/lipics-v2016-authors/jim/sequoidal.tex
A whole bunch of things.

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Sequoidal Categories and Transfinite Games: A Coagebraic Approach to Stateful Objects in Game Semantics\footnote{This work was partially supported by UK EPSRC Grant EP/K037633/1}}
\titlerunning{Sequoidal Categories and Transfinite Games} %optional, in case that the title is too long; the running title should fit into the top page column

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{}
\EventNoEds{2}
\EventLongTitle{7th Conference on Algebra and Coalgebra in Computer Science (CALCO 2017)}
\EventShortTitle{CALCO 2017}
\EventAcronym{CALCO}
\EventYear{2017}
\EventDate{June 12--16, 2017}
\EventLocation{Ljubljana, Slovenia}
\EventLogo{}
\SeriesVolume{xx}
\ArticleNo{xx}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%% Theorems %%
\theoremstyle{plain}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{exercise}{Exercise}
\theoremstyle{definition}
\newtheorem{notation}[theorem]{Notation}

%% Other macros
\newcommand*\from{\colon}
\def \inv {^{-1}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\tensor}{\otimes}
\newcommand{\sequoid}{\oslash}
\renewcommand{\implies}{\multimap}
\newcommand{\comp}[2]{#2 ; #1}
\newcommand{\cprd}{\sqcup}
\newcommand{\bigcprd}{\bigsqcup}
\newcommand{\C}{\mathcal C}
\newcommand{\F}{\mathcal F}
\newcommand{\G}{\mathcal G}
\newcommand{\W}{\mathcal W}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\OP}{\{O,P\}}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\pprefix}{\sqsubsetneqq}
\newcommand{\ppprefix}{\sqsubset}
\DeclareMathOperator{\assoc}{assoc}
\DeclareMathOperator{\lunit}{lunit}
\DeclareMathOperator{\runit}{runit}
\DeclareMathOperator{\sym}{sym}
\newcommand{\braid}{\sym}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\der}{{\mathtt{der}}}
\newcommand{\mult}{{\mathtt{mult}}}
\DeclareMathOperator{\wk}{wk}
\newcommand{\toisom}{{\xrightarrow{\cong}}}
\newcommand{\passoc}{{\mathtt{passoc}}}
\newcommand{\pcomm}{{\mathtt{pcomm}}}
\newcommand{\run}{{\mathtt{r}}}
\newcommand{\lun}{{\mathtt{l}}}
\newcommand{\fcoal}[1]{{\leftmoon #1 \rightmoon}}
\renewcommand{\subset}{\subseteq}
\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\s}{\mathfrak{s}}
\newcommand{\dist}{{\mathtt{dist}}}
\newcommand{\dec}{{\mathtt{dec}}}
\renewcommand{\int}{{\mathtt{coh}}}
\renewcommand{\coh}{\int}
\DeclareMathOperator{\CCom}{CCom}
\newcommand{\catname}[1]{{\mathsf{#1}}}
\newcommand{\Set}{\catname{Set}}
\newcommand{\Ord}{\catname{Ord}}
\newcommand\oppcat[1]{#1^{\mathrm{op}}}
\newcommand{\bN}{{\mathbb{N}}}
\newcommand{\OK}{{\mathtt{OK}}}
\newcommand{\zero}{{\mathtt{zero}}}
\renewcommand{\succ}{{\mathtt{succ}}}
\newcommand{\Var}{{\mathtt{Var}}}
\renewcommand{\read}{{\mathtt{read}}}
\newcommand{\wwrite}{{\mathtt{write}}}
\newcommand{\Rel}{\catname{Rel}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{{\Na}}{\bN}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\fix}{{\mathsf{fix}}}
\newcommand{\eq}{{\mathsf{eq}}}
\newcommand{\co}{{\textrm{\textexclamdown}}}

%%OUR

@@

\begin{abstract}
The non-commutative sequoid operator $\sequoid$ on games was introduced  to capture algebraically the presence of state in history-sensitive strategies in game semantics, by imposing a causality relation on the  tensor product of games.  Coalgebras for the  functor  $A\sequoid\blank$  --- i.e. morphisms from $S$ to $A \sequoid S$ --- may be viewed as state transformers:  if $A\sequoid\blank$ has a \emph{final coalgebra}, $\oc A$, then the anamorphism of such a state transformer encapsulates its explicit state, so that it is shared only between successive invocations.

We study the  conditions under which a final coalgebra $\oc A$ for  $A\sequoid\blank$ is the carrier of a \emph{cofree commutative comonoid} on $A$. That is, it is a model of the exponential of linear logic in which we can construct imperative objects such as reference cells  coalgebraically, in a game semantics setting. We show that if the tensor \emph{decomposes} into the sequoid,  the final coalgebra $!A$ may be endowed with the structure of the cofree commutative comonoid if the natural isomorphism $\oc (A \times B) \cong \oc A \tensor \oc B$ holds. This condition is always satisfied if $\oc A$ is the \emph{bifree algebra} for $A\sequoid\blank$, but in general it is necessary to impose it, as we establish by giving an example of a sequoidally decomposable category of games in which plays will be allowed to have transfinite length. In this category, the final coalgebra for the functor $A\sequoid\blank$ is not the cofree commutative comonoid over $A$: we illustrate this by explicitly contrasting the final sequence for the functor $A\sequoid\blank$ with the chain of symmetric tensor powers used in the construction of the cofree commutative comonoid as a limit by  Melli\'es, Tabareau and Tasson.

\end{abstract}

\section{Introduction}
Game semantics has been used to define a variety of models of higher-order programming languages with  mutable state, including Idealized Algol  \cite{SamsonGuyIAPassive}, and various fragments of ML \cite{AMV,AHM}. Unlike traditional denotational semantics, which typically represent imperative programs as state transformers, the state in these models is completely implicit: local declaration of mutable variables is interpreted as composition with a ``history sensitive'' strategy representing a reference cell. 
This is conceptually simple in principle but leads to some quite combinatorial definitions; a more explicit representation of the current state can be very useful for constructing and reasoning about imperative objects. 
In this paper, we use final coalgebras to forge a link between the explicit and implicit representation of state in game semantics. We study the conditions under which the \emph{cofree commutative comonoid} in categories of games may be derived from a final coalgebra, relating their underlying structure as models of linear logic to classical coalgebraic techniques for describing state.

\subsection{Defining Higher-order Stateful Objects, Coalgebraically}
Let us first motivate the study of the coalgebraically derived cofree comonoid in game semantics by considering a similar but simpler and more familiar phenomenon. 
A \emph{state-transformer} in a symmetric monoidal category is a morphism $f:A \otimes S \rightarrow B \otimes S$ taking an argument together with an input state to a result together with an output state. A well-studied \cite{jacobs} technique in semantics is to  use an appropriate final coalgebra to \emph{encapsulate} the state in such a transformer, allowing multiple  successive invocations, each of which passes its output state as an input state to the next invocation.

For example, consider the category $\Rel$ of sets and relations, with symmetric monoidal structure given by the cartesian product (with unit $I$, the singleton set $\{*\}$). This has finite (bi)products (disjoint unions) so we may define the functor $F(A,S) = (A \otimes S) \oplus I$. For any object (set) $A$, let $A^*$ be the set of finite sequences of elements of $A$ (i.e. the carrier of the free monoid on $A$), and $\alpha:A^* \rightarrow F(A,A^*)$ be the morphism $\{(\varepsilon,\inr(*)\}\cup  \{(aw,(\inl(a,w))\ |\ a \in A,w \in A^*\}$. It is straightforward to show that: 
\begin{lemma}$(A^*,\alpha_A)$ is the final coalgebra  for $F(A,\_)$.
\end{lemma}
Since we have a natural transformation  $\inl_{A,S}:A \otimes S \rightarrow F(A,S)$, we may encapsulate the state in the  state transformer $f:S \rightarrow A \otimes S$ by taking the \emph{anamorphism} of $f;\inl_{A,B}:S \rightarrow F(A,S)$, --- i.e. the unique $F(A,\_)$-coalgebra morphism from $(S,f;\inl)$ into $(A^*,\alpha_{A})$. This  is  a morphism from an initial state $S$ into $A^*$: by definition, composing it with $\alpha:A \rightarrow F(A,A^*)$ (which we can think of as  \emph{invoking} our stateful object) returns a copy of $f$ and together with the encapsulated morphism with updated internal state.

 Distributivity of $\oplus$ over $\otimes$ implies that $F(A \oplus A',S) \cong F(A,S) \oplus F(A',S)$.  This allows state transformers to be aggregated, to construct stateful objects compounded of a series of methods which share access to a common state. 
For example, we may represent a reference cell storing integer values as a state transformer $\cell:\Na \rightarrow (\Na \oplus \Na) \otimes \Na$, obtained by aggregating two ``methods'' which share access to a value in $\Na$ representing the contensts of the cell --- returning a ``read'' of the input state (and leaving it unchanged) or accepting a ``write'' of a new value and using it to update the state.   Thus (with appropriate tagging) it is the relation $\{(i, (\read(i),i))\ |\ i \in \Na\} \cup \{(i,{\mathtt{write}}(j),j)\}$.
The anamorphism of the coalgebra $\cell;\inl:\Na \rightarrow F(\Na \oplus \Na,\Na) $  is the relation from $\Na$ to $(\Na \oplus \Na)^*$ consisting of pairs of the form $(i_1,\read(i_1)^*{\mathtt{write}}(i_2)\read(i_2)^*\ldots)$. Composition with this morphism is precisely the interpretation of new variable declaration in the semantics  in $\Rel$ of the prototypical functional-imperative language \emph{Syntactic Control of Interference} (SCI) given in \cite{Mcsci}.

Coalgebraic methods thus give us a recipe for constructing and using categorical definitions of stateful semantic objects.  In order to fully exploit these, however, we endow $A^*$ with the structure of a \emph{comonoid} in our symmetric monoidal category, by defining morphisms $\delta_A:A^* \rightarrow A^* \otimes A^* = \{(u\cdot v,(u,v))\ |\ u,v \in A^*\}$ and $\epsilon:A \rightarrow I = \{(\varepsilon,*)\}$. In fact, this is the \emph{cofree comonoid} on $A$ --- there is a morphism $\eta_A:A^* \rightarrow A = \{(a,a)\ | \ a \in A\}$ such that for any comonoid $(B,\delta_B,\epsilon_B)$, composition with $\eta_A$ defines an equivalence (natural in $B$)  between the morphisms from $B$ into $A$, and the comonoid morphisms from $(B,\delta_B,\epsilon_B)$ into  $(A^*,\delta_A,\epsilon_A)$.   
\begin{proposition}$(A^*,\delta,\epsilon)$ is the cofree comonoid on $\Rel$.\footnote{The definitions of $\delta$ and $\epsilon$, and the proof that this is the cofree comonoid  may be derived from the fact that $(A^*,\alpha_A)$ is a \emph{bifree algebra} for $F(\_, A)$  --- i.e. $(A^*,\alpha^{-1})$ is an initial algebra for $F(A,\_)$ ($\alpha$ must be an isomorphism by Lambek's lemma). We leave this as an exercise.} 
\end{proposition}
This structure can be used to interpret procedures which share access to a stateful resource such as a reference cell.      
Its main limitation is that we have not defined a \emph{commutative} comonoid  for any non-empty set $A$ (evidently, $\delta$ is not invariant under postcomposition with the symmetry isomorphism of the tensor). Thus we can only model procedures with shared access to the same stateful object if the order in which they are permitted to access it is fixed. (This is precisely the situation in SCI, where the typing system allows sharing  across sequential composition, but not between functions and their arguments.) In order to model sharing of state without this constraint (and build a Cartesian closed category), we need to endow our final coalgebra with the structure of a \emph{cofree commutative comonoid}, proposed as the basis of a model of linear logic by Lafont \cite{LafontCofCommCom}.  The category of sets and relations  does not allow this (the cofree commutative comonoid on an object $A$ in $\Rel$ is given by the set of finite multisets of $A$, which is not a final coalgebra). Hence, we turn to the richer structures of game semantics.
 \subsection{The cofree commutative comonoid as a final colagebra}
We now outline the remainder of the paper. Our main contribution is an investigation of the circumstances in which the  cofree commutative comonoid on $A$ arises from a final coalgebra for the functor $A\sequoid\blank$, where $\sequoid$ (the sequoid) is a non-commutative operation on games  introduced by one of the authors \cite{laird02}.\footnote{We will focus on a particular category of ``history sensitive'', Abramsky-Jagadeesan style games \cite{abramskyjagadeesangames}, but sequoidal structure is a unifying feature of sequential, history-sensitive games.} 
 In this setting, we  can model a state transformer for a program as a morphism $S \to A \sequoid S$ --- i.e., a coalgebra for the functor $A\sequoid\blank$. The final coalgebra for this functor is the exponential game $!A$ introduced by Hyland \cite{hyland1997games}, which corresponds to a $\omega$-fold sequence $A \sequoid (A \sequoid (A \sequoid \ldots ))$; under appropriate conditions, it is the carrier for the  cofree  commutative comonoid on $A$. We aim to characterize these conditions using just the categorical structure,  in order to capture a general class of models and to derive formal principles for coinductively proving  program equivalences. In a nutshell, we require that a certain natural morphism $\oc A \tensor \oc B \to \oc (A\times B)$ is an isomorphism. This can be used to show that $\oc\blank$ gives rise to a strong monoidal functor. Perhaps more surprisingly, it is sufficient to show that $\oc A$ is the cofree commutative comonoid.

This \emph{strong monoidal hypothesis} holds whenever $\oc A$ is a bifree algebra for $A\sequoid\blank$. But we are also interested in cases where $\oc A$ is not bifree --- for example, in categories of ``win games'' and \emph{winning strategies}  \cite{hyland1997games}, which lack the partial maps which can be shown to arise in the bifree case. 
To show that the strong monoidal hypothesis is necessary in general, we introduce a  sequoidal category of games with \emph{transfinite} plays in which it does not hold: because a transfinite interleaving of two sequences of length $\omega$ may have length greater than $\omega$, the final coalgebra for $A\sequoid\blank$  (corresponding  to only $\omega$-many copies of the game $A$) cannot be the carrier for the cofree commutative comonoid.

We compare the  coalgebraic construction of the cofree exponential to the explicit characterization of the latter given by Melli\'es, Tabareau and Tasson \cite{MelliesCofCommCom} as the limit of a chain of symmetric tensor powers. This chain exists in any decomposable sequoidal category: where its limit exists and is preserved by the tensor (the conditions required in \cite{MelliesCofCommCom}) it must be the final coalgebra for $A \sequoid\blank$. However, in our categories of transfinite games, and win games and winning strategies (which may be viewed as games of length $\omega+1$), the construction fails --- this limit is not the cofree commutative comonoid.

@@

\subsection{Game semantics and the sequoidal operator}
We shall present a form of game semantics in the style of \cite{hyland1997games} and \cite{abramskyjagadeesangames}.  A game $A$ is given by a forest (i.e., a prefix-closed set) $P_A$ of \emph{positions}, which are alternating sequences of $O$-moves and $P$-moves.  We call a position a \emph{$P$-position} if it ends with a $P$-move and an \emph{$O$-position} if it ends with an $O$-move.  A \emph{strategy} for a game $A$ is a subforest $\sigma$ of $P_A$ that is closed under $O$-replies to $P$-positions and which satisfies \emph{determinism}: if $sa,sb\in\sigma$, where $s$ is a $P$-position, then $a=b$.  We will adopt the rule that all positions must start with an $O$-move.

We build connectives on games in the usual way.

\begin{description}
  \item[Product] If $(A_i\suchthat i\in I$ is a collection of negative games, then we write $\prod_{i\in I}A_i$ for the game in which player $O$, on his first move, may play in any of the games $A_i$.  From then on, play continues in $A_i$.  If $A_1,A_2$ are games, we write $A_1\times A_2$ for $\prod_{i=1}^2 A_i$.  
  \item[Tensor Product] If $A,B$ are negative games,the tensor product $A\tensor B$ is played by playing the games $A$ and $B$ in parallel, where player $O$ may elect to switch games whenever it is his turn and continue play in the game he has switched to.
  \item[Linear implication] The implication $A\implies B$ is played by playing the game $B$ in parallel with the \emph{negation} of $A$ - that is, the game formed by switching the roles of players $P$ and $O$ in $A$.  Since play in the negation of $A$ starts with a $P$-move, player $O$ is forced to make his first move in the game $B$.  Thereafter, player $P$ may switch games whenever it is her turn.
\end{description}

It is well known (see \cite{abramskyjagadeesangames}, for example) that we may compose strategies $\sigma$ for $A\implies B$ and $\tau$ for $B\implies C$ to get a morphism $\comp\tau\sigma$ for $A\implies C$ and that this structure gives rise to a monoidal category where objects are games, morphisms from $A$ to $B$ are strategies for $A\implies B$ and the tensor product and linear implication are given by $A\tensor B$ and $A\implies B$.  We call this category $\G$.  $\G$ has all products, given by $\prod_{i\in I}A_i$ as above.

@@

The sequoid does not give rise to a functor $\G\times\G\to \G$ in the way that the tensor product does: playing strategies $\sigma$ for $A\implies B$ and $\tau$ for $C \implies D$ in parallel does not necessarily give rise to a valid strategy for $(A\sequoid C)\implies (B \sequoid D)$, since player $O$ might end up playing in $C$ before anyone has played in $A$.  However, if we require that the strategy $\sigma$ is \emph{strict} -- that is, player $P$'s reply to the opening move in $B$ is always a move in $A$ -- then we do get a valid strategy $\sigma\sequoid\tau$ for $(A\sequoid C)\implies (B\sequoid D)$ and, moreover, $\sigma\sequoid\tau$ is strict.  If we write $\G_s$ for the category of games with \emph{strict} strategies as morphisms, then we see that $\blank\sequoid\blank$ gives us a functor $\G_s\times \G\to\G_s$.  In fact, the isomorphism $\passoc$ tells us that this functor is a right monoidal category action of $\G$ upon $\G_s$.

>>>>

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Sequoidal Categories and Transfinite Games: Towards a Coalgebraic Approach to Linear Logic\footnote{This work was partially supported by [TODO: find the appropriate grant codes]}}
\titlerunning{Sequoidal Categories and Transfinite Games} %optional, in case that the title is too long; the running title should fit into the top page column

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%% Theorems %%
\theoremstyle{plain}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{notation}[theorem]{Notation}

%% Other macros
\newcommand*\from{\colon}
\def \inv {^{-1}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\Rel}{{\mathsf{Rel}}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{{\Na}}{{\mathbb{N}}}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\tr}{\oslash}
\newcommand{\fix}{{\mathsf{fix}}}
\newcommand{\eq}{{\mathsf{eq}}}
%\\write{{\mathtt{write}}}
\def\read{{\mathtt{read}}}
\def\({\ensuremath{( \! [}}
\def\){\ensuremath{] \! )}}
\newcommand{\perm}{{\mathsf{perm}}}
\newcommand{\tensor}{\otimes}
\newcommand{\sequoid}{\oslash}
\renewcommand{\implies}{\multimap}
\newcommand{\comp}[2]{#1 \circ #2}
\newcommand{\cprd}{\sqcup}
\newcommand{\bigcprd}{\bigsqcup}
\newcommand{\C}{\mathcal C}
\newcommand{\F}{\mathcal F}
\newcommand{\G}{\mathcal G}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\OP}{\{O,P\}}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\pprefix}{\sqsubsetneqq}
\newcommand{\ppprefix}{\sqsubset}
\DeclareMathOperator{\assoc}{assoc}
\DeclareMathOperator{\lunit}{lunit}
\DeclareMathOperator{\runit}{runit}
\DeclareMathOperator{\sym}{sym}
\newcommand{\braid}{\sym}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\der}{{\mathtt{der}}}
\newcommand{\mult}{{\mathtt{mult}}}
\DeclareMathOperator{\wk}{wk}
\newcommand{\toisom}{{\xrightarrow{\cong}}}
\newcommand{\passoc}{{\mathtt{passoc}}}
\newcommand{\pcomm}{{\mathtt{pcomm}}}
\newcommand{\run}{{\mathtt{r}}}
\newcommand{\lun}{{\mathtt{l}}}
\newcommand{\fcoal}[1]{{\leftmoon #1 \rightmoon}}
\renewcommand{\subset}{\subseteq}
\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\s}{\mathfrak{s}}
\newcommand{\dist}{{\mathtt{dist}}}
\newcommand{\dec}{{\mathtt{dec}}}
\renewcommand{\int}{{\mathtt{int}}}
\DeclareMathOperator{\CCom}{CCom}
\newcommand{\catname}[1]{{\normalfont\textbf{#1}}}
\newcommand{\Set}{\catname{Set}}
\newcommand{\Ord}{\catname{Ord}}
\newcommand\oppcat[1]{#1^{\mathrm{op}}}

@@

\begin{abstract}
  In \cite{laird02}, Laird introduces the concept of a \emph{sequoidal category} as a formalization of causality in game semantics.  A sequoidal category is like a monoidal category with an extra connective, $\sequoid$, that allows one to construct an exponential object as a final coalgebra \cite{martinsthesis},  Under a further hypothesis, it is possible to show that this final coalgebra allows one to construct cofree commutative comonoids in the category, giving us a model of the exponential connective $\oc$ from linear logic.  In the first part of this note, we review the coalgebraic arguments for constructing the cofree commutative comonoid, which are known but do not yet appear in print.  In the second part, we show that the extra hypotheses are necessary by outlining a definition of \emph{transfinite game}, in which the carriers of the sequoidal coalgebra and the cofree commutative comonoid do not coincide.
\end{abstract}

\section{Defining Higher-order Stateful Objects, Coalgebraically}
In this section we motivate the study of the coalgebraically derived cofree comonoid in game semantics by considering a similar but simpler and more familiar phenomenon. A \emph{state-transformer} in a symmetric monoidal category is a morphism $f:A \otimes S \rightarrow B \otimes S$ taking an argument together with an input state to a result together with a input state. A well-studied \cite{jacobs} technique in semantics is to  use an appropriate final coalgebra to \emph{encapsulate} the state in such a transformer, allowing multiple  successive invocations, each of which passes its output state as an input state to the next invocation.

For example, consider the category $\Rel$ of sets and relations, with symmetric monoidal structure given by the cartesian product (with unit $I$, the singleton set $\{*\}$). This has finite (bi)products given by the disjoint union of sets: define the functor $F(A,S) = (A \otimes S) \oplus I$

For any object (set) $A$, let $A^*$ be the set $A^*$ of finite sequences of elements of $A$ (i.e. the carrier of the free monoid on $A$), and $\alpha:A^* \rightarrow F(A,A^*)$ be the morphism $\{(\varepsilon,\inr(*)\}\cup  \{(aw,(\inl(a,w))\ |\ a \in A,w \in A^*\}$   
\begin{lemma}$(A^*,\alpha_A)$ is a final coalgebra  for $F(A,\_)$.
\end{lemma}
\begin{proof}$\Rel$ may be cpo-enriched with the inclusion order: $A^*$ is a \emph{minimal invariant} for $F(A,\_)$ with respect to this order.
\end{proof}
Since we have a natural transformation (left injection) $\inl_{A,S}:A \otimes S \rightarrow F(A,S)$, we may encapsulate the state in the  state transformer $f:S \rightarrow A \otimes S$ by taking the \emph{anamorphism} of $f;\inl_{A,B}:S \rightarrow F(A,S)$, --- i.e. the unique $F(A,\_)$-coalgebra morphism from $(S,f;\inl)$ into $(A^*,\alpha_{A})$. This  is  a morphism from an initial state $S$ into $A^*$: by definition, composing it with $\alpha:A \rightarrow F(A,A)$ (which we can think of as  \emph{invoking} our stateful object) returns a copy of $f$ and uses it to update the internal state.

By distributivity of $\oplus$ over  $\otimes$ we have a natural transformation $t: F(A,S) \otimes S' \rightarrow F(A,S \otimes S')$. We may use this to encapsulate  state transformers of general form: given $f:A \otimes S \rightarrow B \otimes S$, taking the anamorphism of the $F(B,\_)$-colagebra:    
$$A^* \otimes S  \stackrel{\alpha_A \otimes S}{\longrightarrow} F(A,A^*) \otimes S \stackrel{t}{\longrightarrow} F(A,A^* \otimes S)$$

 Distributivity of $\oplus$ over $\otimes$ also implies that $F(A \oplus A',S) \cong F(A,S) \oplus F(A',S)$.  This allows state transformers to be aggregated, to construct stateful objects compounded of a series of methods which share access to a common state.
i.e. given morphisms $f_1:A_1 \otimes S \rightarrow B_1 \otimes S, \ldots , f_n:A_n \otimes S \rightarrow B_n \otimes S$, encapsulating  $\dist;(f_1 \oplus \ldots \oplus f_n);\dist^{-1}$, gives a morphism from $(A_1 \oplus \ldots \oplus A_n)^* \otimes S$ into $(B_1 \oplus \ldots \oplus B_n)^*$.

For example, we may represent a reference cell storing integer values as a state transformer $\cell:\Na \rightarrow (\Na \oplus \Na) \otimes \Na$, obtained by aggregating two ``methods'' which share access to a state consisting of a single integer, representing the contents of the cell --- returning a ``read'' of the input state (and leaving it unchanged) or accepting a ``write'' of a new value and using it to update the state.   Thus (with appropriate tagging) it is the relation $\{(i, (\read(i),i))\ |\ i \in \Na\} \cup \{(i,{\mathtt{write}}(j),j)\}$.
The anamorphism of the coalgebra $\cell;\inl:\Na \rightarrow F(\Na \oplus \Na,\Na) $  is the relation from $\Na$ to $(\Na \oplus \Na)^*$ consisting of pairs of the form $(i_1,\read(i_1)^*{\mathtt{write}}(i_2)\read(i_2)^*\ldots)$ Composition with this morphism is precisely the denotation of new variable declaration in the semantics of  Reynolds \emph{Syntactic Control of Interference} (SCI) in $\Rel$ given in \cite{Mcsci}.

Coalgebraic methods thus give us a recipe for constructing and using categorical definitions of stateful semantic objects, avoiding direct definitions which are rather combinatorial to work with.  In order to fully exploit these, however, we may endow $A^*$ with the structure of a \emph{comonoid} by defining morphisms $\delta_A:A^* \rightarrow A^* \otimes A^* = \{(u\cdot v,(u,v))\ |\ u,v \in A^*\}$ and $\epsilon:A \rightarrow I = \{(\varepsilon,*)\}$

 In fact, this is the \emph{cofree comonoid} on $A$ --- there is a morphism $\eta_A:A^* \rightarrow A = \{(a,a)\ | \ a \in A\}$ such that for any comonoid $(B,\delta_B,\epsilon_B)$, composition with $\eta_A$ defines an equivalence (natural in $B$)  between the morphisms from $B$ into $A$, and the comonoid morphisms from $(B,\delta_B,\epsilon_B)$ into  $(A^*,\delta_A,\epsilon_A)$.   
\begin{proposition}$(A^*,\delta,\epsilon)$ is the cofree comonoid on $\Rel$.
\end{proposition}
The definitions of $\delta$ and $\epsilon$, and the proof that this is the cofree comonoid  may be derived from the fact that $(A^*,\alpha_A)$ is a \emph{bifree algebra} for $F(\_, A)$  --- i.e. $(A^*,\alpha^{-1})$ is an initial algebra for $F(A,\_)$ ($\alpha$ must be an isomorphism by Lambek's lemma). We leave this as an exercise.

This structure can be used to interpret procedures which share access to a stateful resource such as a reference cell, the creation of multiple copies of such procedures ($(\_)^*$ is a monoidal comonad)  etc.      
Its main limitation is that we have not defined a \emph{commutative} monoid --- evidently it is not the case that $\delta_A;\sym_{A,A} = \delta_A$ (where $\sym_{A,B}:A \otimes B \rightarrow B \otimes A$ is the symmetry isomorphism for the tensor) for any non-empty set $A$. Thus we can only model procedures with shared access to the same stateful object if the order in which they are permitted to access it is predetermined. This is precisely the situation in SCI, where the typing system allows sharing between across sequential composition, but not between functions and their arguments. In order to model sharing of state between functions without this constraint (and build a Cartesian closed category), we need to endow our final coalgebra with the structure of a cofree \emph{commutative} comonoid.  
The category of sets and relations  does not allow this (the cofree commutative comonoid on an object $A$ in $\Rel$ is given by the set of finite multisets of $A$, which is not a final coalgebra). By properly reflecting the interleaving between function calls, game semantics does support a coalgebraic cofree commutative comonoid, leading to models of stateful higher-order languages such as Idealized Algol and core ML. As we shall show by example and counterexample, the richer structure of games also means that the relationship between initial and final coalgebras, and cofree objects is more subtle  (note, for example, that any final coalgebra in $\Rel$ must be bifree, by its self-duality).

@@

\subsection{Game semantics and the sequoidal operator}

We shall present a form of game semantics in the style of \cite{hyland1997games} and \cite{abramskyjagadeesangames}.  A game will be given by a tuple
\[
  A = (M_A, \lambda_A, b_A, P_A)
  \]
where
\begin{itemize}
  \item $M_A$ is a set of moves.
  \item $\lambda_A\from M_A\to\OP$ is a function designating each move as either an \emph{$O$-move} or a \emph{$P$-move}.
  \item $b_A\in\OP$ is a choice of starting player.
  \item $P_A\subset M_A^*$ is a prefix-closed set of alternating plays (so if $sab\in P_A$ then $\lambda_A(a)=\neg\lambda_A(b)$) such that if $as\in P_A$ then $\lambda_A(a)=b_A$.
\end{itemize}

We call $sa\in P_A$ a \emph{$P$-position} if $a$ is a $P$-move and an \emph{$O$-position} if $a$ is an $O$-move.

A \emph{strategy} for player $P$ for a game $A$ is identified with the set of positions that may arise when playing according to that strategy.  Namely, it is a non-empty prefix-closed subset $\sigma\subset P_A$ satisfying the two conditions:
\begin{description}
  \item[(sO)] If $s\in\sigma$ is a $P$-position and $a$ is an $O$-move such that $sa\in P_A$, then $sa\in\sigma$.
  \item[(sP)] If $sa,sb\in\sigma$ are $P$-positions, then $a=b$.
\end{description}

We shall now concentrate on games $A$ for which $b_A=O$, called \emph{negative games}.  We shall informally describe the standard connectives on negative games:

\begin{description}
  \item[Product] If $A$ and $B$ are negative games then the \emph{product} $A\times B$ is the game given by placing the game trees for $A$ and $B$ side by side: that is, player $O$ may play his first move either in $A$ or in $B$.  Thereafter, play continues in the game that player $O$ has chosen.
  \item[Tensor Product] The tensor product $A\tensor B$ is also played by playing the games $A$ and $B$ in parallel, but this time player $O$ may elect to switch games whenever it is his turn and continue play in the game he has switched to.
  \item[Linear implication] The implication $A\implies B$ is played by playing the game $B$ in parallel with the \emph{negation} of $A$ - that is, the game formed by switching the roles of players $P$ and $O$ in $A$.  Since play in the negation of $A$ starts with a $P$-move, player $O$ is forced to make his first move in the game $B$.  Thereafter, player $P$ may switch games whenever it is her turn.
\end{description}

If $A,B,C$ are negative games, $\sigma$ is a strategy for $A\implies B$ and $\tau$ is a strategy for $B\implies C$, then we may form a strategy $\comp\tau\sigma$ for $A\implies C$ by setting
\[
  \sigma\|\tau = \{\s\in (M_A\cprd M_B\cprd M_C)^*\suchthat \s\vert_{A,B}\in\sigma,\;s\vert_{B,C}\in\tau\}
  \]
and then defining
\[
  \comp\tau\sigma = \{\s\vert_{A,C}\suchthat\s\in\comp\tau\sigma\}
  \]
It is well known (see, for example, \cite{abramskyjagadeesangames}) that $\comp\tau\sigma$ is indeed a strategy for $A\implies C$ and that this form of composition is associative and has an identity.  It is also well known that the resulting category $\G$ of games and strategies has products given by the operator $\times$ and a symmetric monoidal closed structure given by the operations $\tensor$ and $\implies$.

@@

One further question to ask is: does the sequoid operator give rise to a functor $\blank\sequoid\blank\from\G\times\G\to\G$, as the tensor operator does?  The answer is no: indeed, let $A,B,C,D$ be negative games, let $\sigma$ be a strategy for $A\implies C$ and let $\tau$ be a strategy for $B\implies D$.  Our aim is to construct a natural strategy $\sigma\sequoid\tau$ for $(A\sequoid B)\implies (C\sequoid D)$.  There is an obvious way to try and do this: player $P$ should play according to the strategy $\sigma$ whenever player $O$'s last move was in $A$ or $C$, and according to $\tau$ whenever player $O$'s last move was in $B$ or $D$.

We show that this does not in general give us a strategy for $(A\sequoid B)\implies (C\sequoid D)$.  Suppose that $\sigma$ is such that player $P$'s response to some opening move in $C$ is another move in $C$ and suppose that $\tau$ is such that player $P$'s response to some opening move in $D$ is a move in $B$ (for example, $\tau$ is a copycat strategy).  Then we end up with the following sequence of events in the game $(A\sequoid B)\implies(C\sequoid D)$  :
\begin{enumerate}
  \item Player $O$ starts with a move in $C$ (as he must).
  \item Player $P$ responds according to $\sigma$ with another move in $C$.
  \item Player $O$ decides to switch games and play a move in $D$.
  \item Player $P$ responds according to $\tau$ with a move in $B$.
\end{enumerate}
